Robust Quantum Optimal Control

### 2020-07-17 paper/paper.tex
progress on paper

%% SECTION 1
\section{Introduction}
(Existing work) The leading model of universal
quantum computation is gate-based. Analytic methods
have been demonstrated to construct gates.
Some focus on solving the TDSE analytically
\cite{}Analytic methods have
been demonstrated to construct gates
\cite{zhang2020universal, huang2020engineering,
  han2020experimental, xu2020nonadiabatic}.
Most methods focus on optimizing a few aspects of the gate, but not all
experimentally relevant aspects. There is a growing literature
on numerical
techniques to construct gates \cite{leung2017speedup,
  goerz2019krotov, doria2011optimal, abdelhafez2019gradient,
  machnes2015gradient}.
%% TODO: Cite CRAB, GOAT, KROTOV, etc.
These methods formulate the quantum optimal
control problem as unconstrained.
They typically rely on zeroth-order or first-order optimizers.
They are not
sofisticated enough to handle all of the relevant
constraints simultaneously.
Although some analytic techniques exist to design pulses robust
to decoherence, no numerical techniques have been presented to
design pulses robust to decoherence (as far as the author is aware).

(This work) We employ the trajectory optimization
literature to formulate the quantum
optimal control problem as a constrained optimization problem.
We study the quantum optimal control problem on the fluxonium.
We outline experimentally realistic constraints and map them to
the trajectory optimization framework. For the device we study
we achieve a 2x increase in $T_{1}$ times.
We present two methods for achieving robustness to system parameter deviations,
and compare to existing dynamic decoupling methods. We find that 
our methods beat dynamic decoupling and mitigate dephasing by order X.

(Outline)  First we formulate the quantum optimal control problem
in the trajectory optimization framework. Then, we introduce the dynamics
of the fluxonium device and outline experimental considerations
relevant to gate construction. Next we outline a method for
making the optimization $T_{1}$ aware. Finally, we present
some methods for engineering robustness to decoherence and
compare them to existing techniques.

%% SECTION 2
\section{QOC + AL-iLQR}
(QOC Problem Statement) I have some initial
configuration and I want to reach some target configuration,
either single- or multi-state transfer, subject to the dynamics
\begin{align}
  \partial_{t} \ket{\psi(t)} &= -\frac{i}{\hbar} H(t) \ket{\psi(t)}
\end{align}
Common numerical techniques include approximating the analytic unitary propagator solution,
or employing explicit/implicit Runge-Kutta methods of the form
\begin{align}
  \ket{\psi^{(k + 1)}} &= \ket{\psi^{(k)}} + \partial_{t} \ket{\psi^{(k)}} \cdot dt
\end{align}
The interesting part is that your hamiltonian $H(t)$ has a time-dependent
control parameter $u(t)$ that the experimentalist gets to control, e.g. flux threading
a superconducting junction.
\begin{align}
  H(t) &= H_{0} + \sum_{i = 1}^{m} a_{m}(u_{m}(t)) H_{m}
\end{align}
Most often $a_{m}(u) = u$ but any arbitrary dependence is allowed.
The technique of adjusting $\ket{\psi_{N}}$ based on $\textbf{u}(t)$
is the goal of the quantum optimal control optimizer. This domain is
know as sensitivity analysis. GRAPE does
\begin{align}
  \min_{U:N-1 \times m} &1 - {\lvert \braket{\psi_{N}(U) \lvert \psi_{t}} \lvert}^{2}
\end{align}
subject to the gradient descent update procedure
\begin{align}
  U^{'} &= U - \alpha \nabla_{U} (1 - {\lvert \braket{\psi_{N}(U) \lvert \psi_{t}} \lvert}^{2})
\end{align}
In general they add more cost functions than target fidelity.
%% TOOD: Other update procedures might differ, e.g. Krotov does some sequential updates
%% and Houck, Google use black box optimizers.

(AL-iLQR Problem Statement) Trajectory optimization gives us gaurantees
about our updates via Ricatti recursion
and allows us to put constraints on our cost functions.
%% This is e.q. 2 from ALTRO paper
\begin{equation}
\begin{aligned}[b]
  \min_{U:N-1 \times m} &\mathcal{L}_{N}(x_{N}, \lambda_{N}, \mu_{N})\\
  &+ \sum_{k = 1}^{N - 1} \mathcal{L}_{k}(x_{k}, u_{k}, \lambda_{k}, \mu_{k})\\
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}[b]
  \mathcal{L}_{k}(x_{k}, &u_{k}, \lambda_{k}, \mu_{k}) =\\
  &(\lambda_{k} + \frac{1}{2}I_{\mu_{k}} c_{k}(x_{k}, u_{k}))^{T} c_{k}(x_{k}, u_{k})\\
  &+ (x_{k} - x_{f})^{T} Q (x_{k} - x_{f}) + u_{k}^{T} R u_{k}\\
\end{aligned}
\end{equation}
The important point is that there is an update step (e.q. 17 from ALTRO paper) where we send $\lambda \rightarrow \infty$
and get all of the nice convergence properties. The weights are adjusted dynamically between iterations until
all of our constraints are satisfied. The Markovian decision structure of the problem allows
us to apply differentiable dynamic programming to garauntee that the update for each control $U_{k}$ is
opimal, as apposed to the greedy updates of first-order optimizers like the naive gradient descent.

@@

%% SECTION 3
\section{Our Problem}
(Fluxonium Device) In the two-level
approximation we have
\begin{align}
  H/h &= \omega_{q} \frac{\sigma_{z}}{2} + A(\Phi_{ext}) \frac{\sigma_{x}}{2}
\end{align}
This approximation is good up to the avoided crossing at $\Phi_{\textrm{ext}} = 0.35 \Phi_{0}$. We get $A$
by converting via $\bra{g} \hat{\phi} \ket{e}$.
We use the gates $X/2, Y/2, Z/2$. $X/2, Y/2$ are universal
up to an arbitrary Z rotation.
%% TODO: list device parameters

@@

%% F4.1
%% TODO: Update images. Use new gates constructed with
%% spline T1 and use smaller sim times.
\begin{figure}[ht]
  \begin{subfigure}{.33\textwidth}
    \includegraphics[width=\linewidth]{assets/00045_spin16.png}
    \label{fig:sub-first}
  \end{subfigure}
  \begin{subfigure}{.33\textwidth}
    \includegraphics[width=\linewidth]{assets/00047_spin16.png}
    \label{fig:sub-second}
  \end{subfigure}
  \begin{subfigure}{.33\textwidth}
    \includegraphics[width=\linewidth]{assets/00049_spin16.png}
    \label{fig:sub-third}
  \end{subfigure}
  \caption{Master equation simulation with $T_{1}$ dissipation for all basis gates.}
  \label{fig:fig}
\end{figure}

>>>>

%% SECTION 1
\section{Introduction}
(Existing work) The leading model of universal
quantum computation is gate-based. The Quantum Optimal Control (QOC)
literature has saught to answer the question "What is the best way to construct a quantum gate?".
Analytic methods
have been demonstrated to construct quantum gates
\cite{zhang2020universal, huang2020engineering, han2020experimental,
  xu2020nonadiabatic, carlini2005quantum}.
Most methods rely on solving the time-dependent
Schroedinger equation analytically or deriving equations of motion
from a suitable Lagrangian.

Some analytical methods
have been presented to address realistic experimental constraints.
Considering the dynamical and geometric phases of state
evolution has led to methods for achieving
robustness to pure dephasing in quantum systems
\cite{xu2020nonadiabitc, han2020experimental, merrill2014progress}.
Additionally, suitable choices of bases have been demonstrated that
allow the experimentalist to infer tradeoffs in longitudinal
and pure dephasing coherence times \cite{huang2020engineering}.
To date, however, no analytical method has been presented
to construct a gate while satisfying all relevant experimental
constraints. Furthermore, most analytic methods have only
been demonstrated in the two-level approximation and become
cumbersome to solve for larger Hilbert space sizes.

Numerical quantum control techniques have developed in parallel
with the analytic frameworks. Numerical quantum control
techniques integrate the time-dependent Schroedinger
equation and use sensitivity analysis to optimize target metrics
of the quantum gate, such as fidelity
\cite{leung2017speedup,  goerz2019krotov, doria2011optimal,
  abdelhafez2019gradient, machnes2015gradient, leng2019robust}.
These techniques typically employ zeroth- or first-order
optimizers and make no garauntees of optimality.
Further, they formulate the quantum control
problem as unconstrained or use
projective gradient methods \cite{machnes2015gradient}.
The latter approach relies on constructing
bases for constraint spaces--which is not always feasible--and
may hinder convergence.

(This work) We employ the trajectory optimization
literature to formulate the quantum
optimal control problem as a constrained optimization problem.
The trajectory optimization framework allows us to handle arbitrary
constraints on the evolution of the quantum state. Additionally,
we introduce methods for constructing gates robust to longitudinal
relaxation and pure dephasing, the dominant barrier to experimentally
realizing practical quantum computing.

We study the quantum optimal control problem on the fluxonium qubit.
We outline experimentally realistic constraints and map them to
the trajectory optimization framework. For the device we study
we achieve a 1.5x increase in $T_{1}$ times.
We present two methods for achieving robustness to pure dephasing,
and compare them to existing dynamic decoupling methods. We find that 
our methods mitigate dephasing by order X over dynamic decoupling.

(Outline)  First we formulate the quantum optimal control problem
in the trajectory optimization framework. Then, we introduce the dynamics
of the fluxonium device and outline experimental considerations
relevant to gate construction. Next we outline a method for
making the optimization $T_{1}$ aware. Finally, we present
some methods for engineering robustness to pure dephasing and
compare them to existing techniques.

%% SECTION 2
\section{QOC + AL-iLQR}
(QOC Problem Statement) Here we introduce the notation
we will use throughout the paper,
review the quantum optimal control problem statement,
and introduce the trajectory optimization framework.
Quantum optimal control concerns the evolution of
a state $\ket{\psi(t)}$ governed by the time-dependent
Schroedinger equation
\begin{align}
  \partial_{t} \ket{\psi(t)} &= -\frac{i}{\hbar} H(\textbf{u}(t), t) \ket{\psi(t)}
\end{align}
The evolution is sometimes cast with $\rho(t) = \ket{\psi(t)}\bra{\psi(t)}$
and the Lindblad master equation to model the decoherence of the
state explicitly. The Hamiltonian
depends arbitrarily on the controls $\textbf{u}(t)$.

Numerical quantum optimal control techniques make
the problem tractable by discretizing the problem into $N$
time steps. Typical integration techniques rely on approximating
the analytical solution to the time-dependent Schroedinger equation
or using explicit integration methods, e.g. Runge-Kutta, of the form
\begin{align}
  \ket{\psi_{k + 1}} \approx \ket{\psi_{k}} + \partial_{t} \ket{\psi_{k}} \cdot \Delta t_{k}
\end{align}

Quantum optimal control seeks the control
parameters that minimize a functional $J(u, \ket{\psi_{N}(u)})$.
In the simplest case the functional
$J = 1 - {\lvert \braket{\psi_{f} \lvert \psi_{N}(u)} \rvert}^{2}$
which is the infidelity between the inital state evolved
to the final time step ($\ket{\psi_{N}(u)}$)
and the target state ($\ket{\psi_{f}}$). In general
$J$ is a linear combinaion of cost functions on $\ket{\psi_{N}(u)}$, e.g.
forbidden-state occupation, as well as
cost functions on $u$, e.g. the norm of the control amplitudes
\cite{leung2017speedup}.

(AL-iLQR Problem Statement) The trajectory optimization
literature solves a more general class of non-linear programs that resemble
the quantum optimal control problem. The quantum optimal control
problem is a specific case of the linear quadratic regulator (LQR).
LQR is so called because the dynamics are linear in the state and
the cost function is quadratic in the state. In the LQR formulation
the same functional is evaluated at each time step
\begin{equation}
  \begin{aligned}
    J_{\textrm{LQR}} = \tilde{x}_{N}^{T} Q_{N} \tilde{x}_{N}
    &+ \sum_{k = 0}^{N - 1} \tilde{x}_{k}^{T} Q_{k} \tilde{x}_{k} + u_{k}^{T} R_{k} u_{k}
  \end{aligned}
\end{equation}
where $\tilde{x}_{k} = x_{k} - x_{f}$ is the difference between the current
and final state and $Q_{k}, R_{k}$ are matrices that define the penalty metric.

The advantage of the LQR formulation
is that there exists a dynamic programming algorithm to compute the
optimal update to the controls $u_{k}$ at each time step.
This algorithm employs the Ricatti recursion (see Appendix).

In order to incorporate constraints into the problem we employ
the augmented lagrangian method. This method adaptively
updates the penalty of constraints
%% LEFT OFF HERE

Trajectory optimization gives us gaurantees
about our updates via Ricatti recursion
and allows us to put constraints on our cost functions.
%% This is e.q. 2 from ALTRO paper
\begin{equation}
\begin{aligned}[b]
  \min_{U:N-1 \times m} &\mathcal{L}_{N}(x_{N}, \lambda_{N}, \mu_{N})\\
  &+ \sum_{k = 1}^{N - 1} \mathcal{L}_{k}(x_{k}, u_{k}, \lambda_{k}, \mu_{k})\\
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}[b]
  \mathcal{L}_{k}(x_{k}, &u_{k}, \lambda_{k}, \mu_{k}) =\\
  &(\lambda_{k} + \frac{1}{2}I_{\mu_{k}} c_{k}(x_{k}, u_{k}))^{T} c_{k}(x_{k}, u_{k})\\
  &+ (x_{k} - x_{f})^{T} Q (x_{k} - x_{f}) + u_{k}^{T} R u_{k}\\
\end{aligned}
\end{equation}
The important point is that there is an update step (e.q. 17 from ALTRO paper) where we send $\lambda \rightarrow \infty$
and get all of the nice convergence properties. The weights are adjusted dynamically between iterations until
all of our constraints are satisfied. The Markovian decision structure of the problem allows
us to apply differentiable dynamic programming to garauntee that the update for each control $U_{k}$ is
opimal, as apposed to the greedy updates of first-order optimizers like the naive gradient descent.

@@

%% SECTION 3
\section{QOC on the Fluxonium}
(Fluxonium Device) In the two-level
approximation we have
\begin{align}
  H/h &= \omega_{q} \frac{\sigma_{z}}{2} + A(\Phi_{ext}) \frac{\sigma_{x}}{2}
\end{align}
This approximation is good up to the avoided crossing at $\Phi_{\textrm{ext}} = 0.35 \Phi_{0}$. We get $A$
by converting via $\bra{g} \hat{\phi} \ket{e}$.
We use the gates $X/2, Y/2, Z/2$. $X/2, Y/2$ are universal
up to an arbitrary Z rotation.
%% TODO: list device parameters

@@

%% F4.1
%% TODO: Update images. Use new gates constructed with
%% spline T1 and use smaller sim times.
\begin{figure}[ht]
  \begin{subfigure}{.33\textwidth}
    %% \includegraphics[width=\linewidth]{assets/00045_spin16.png}
    \label{fig:sub-first}
  \end{subfigure}
  \begin{subfigure}{.33\textwidth}
    %% \includegraphics[width=\linewidth]{assets/00047_spin16.png}
    \label{fig:sub-second}
  \end{subfigure}
  \begin{subfigure}{.33\textwidth}
    %% \includegraphics[width=\linewidth]{assets/00049_spin16.png}
    \label{fig:sub-third}
  \end{subfigure}
  \caption{Master equation simulation with $T_{1}$ dissipation for all basis gates.}
  \label{fig:fig}
\end{figure}
